\documentclass{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage[margin=1in]{geometry}

\title{Zero-inflated Dirichlet}
\date{}

\begin{document}
\maketitle 
The main issue with a novelty detection using density estimation as an approach 
(even in $\mathbb{S}_p^{d-1}$ is density estimation for edge cases 
(observations with values near 0) causing density evaluation to diverge.  I \emph{want} a
Dirichlet-like distribution that allows for a point-mass at 0, in presumably all but 1 dimension.

I had thought that one means of dealing with that would be a \emph{zero-inflated} projected 
Gamma. I thought I'd try it for Dirichlet first, as its Jacobian is easier to handle.
(evaluates to $r^{d-1}$).

Given the construction of Dirichlet as a normalization of Gamma RV's to the unit simplex,
then we start with a zero-inflated Dirichlet,
\[
f(x\mid\pi,\alpha) \sim (1 - \pi)\delta_{(0)}(x) + \pi\mathcal{G}(x\mid\alpha).
\]
But we need it in product form, rather than sum form, so
\[
f(x\mid\alpha) = \int_{\nu}\delta_{(0)}(x)^{1 - \nu}\mathcal{G}(x\mid\alpha,1)^{\nu}\text{d}G(\nu).
\]
Let $\bm{Y} \in \mathbb{S}_1^{d-1} \implies \sum_{\ell = 1}^d Y_{\ell} = 1$, $Y_{\ell} \in [0, 1] \forall \ell$.  Then, the distribution of $\bm{y}\mid\bm{\alpha} = \int_r \int_{\nu} \prod_{\ell = 1}^d f(ry_{\ell}\mid\alpha_{\ell},\nu_{\ell})\text{d}G(\nu)\text{d}r$.
\[
\begin{aligned}
f(\bm{y}\mid\bm{\alpha}) &= \int_r \int_{\nu} \prod_{\ell = 1}^d \delta_{(0)}(ry_{i\ell})^{1 - \nu_{\ell}}\mathcal{G}(ry_{\ell}\mid\alpha_{\ell},1)^{\nu_{\ell}}r^{d-1}\text{d}G(\nu)\text{r} \\ 
&= \int_r\int_{\nu}\prod_{\ell = 1}^d\delta_{(0)}(y_{i\ell})^{1 - \nu_{\ell}}\left(\frac{(ry_{\ell})^{\alpha_{\ell} - 1} e^{-ry_{ell}}}{\Gamma(\alpha_{\ell})}\right)^{\nu_{\ell}}\text{d}G(\nu)\text{d}r
\end{aligned}
\]
Note the omission of $r$ from the Dirac delta in the second line; $r$ will never be 0, so the Dirac delta function is the same either way.  Then this breaks apart as:
\[
f(\bm{y}\mid\bm{\alpha}) = \int_{r}\int_{\nu}\prod_{\ell = 1}^{d}\delta_{(0)}(y_{i\ell})^{1 - \nu_{\ell}}\left(\frac{y_{\ell}^{\alpha_{\ell} - 1} e^{-y_{ell}}}{\Gamma(\alpha_{\ell})}\right)^{\nu_{\ell}}\times r^{\sum_{\ell = 1}^d\nu_\ell(\alpha_{\ell} - 1)}e^{-r\sum_{\ell = 1}^dy_{\ell}}r^{d-1}\text{d}G(\nu)\text{d}r
\]
Pulling the full conditional for $r$ from the above equation, we get:
\[
\begin{aligned}
f(r\mid\bm{\alpha},\nu) &\propto 
    r^{\sum_{\ell = 1}^d\nu_\ell(\alpha_{\ell} - 1)}
    e^{-r\sum_{\ell = 1}^dy_{\ell}}
    r^{d-1}
    \text{d}r\\
    &= r^{\sum_{\ell = 1}^d\nu_\ell\alpha_{\ell} - \sum_{\ell - 1}^{d}\nu_{\ell} + d - 1}\times e^{-r\sum_{\ell = 1}^dy_{\ell}}
\end{aligned}
\]
The problem arises here.  In Dirichlet, $d$ in the Jacobian is cancelled out by $f(r\bm{y}\mid\alpha)$  Here, we're only cancelling out $\sum_{\ell = 1}^d\nu_{\ell}$ of that $d$.
That'll be a valid gamma RV, so we can integrate it out from the above, but I'm thinking that that mismatch is going to cause a sampler to diverge.

Do you see a way out?
    
\end{document}

% EOF